---
title: "Join health, pop and environmental data for Repository"
author: "Julia Barreto"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    thumbnails: false
    lightbox: true
    gallery: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = F, message = F, warning = F, error= F)
library(here)
library(sf)
library(tidyverse)
library(kableExtra)
library(readxl)
library(tibble)
```

Load data from raw health datasets for each country:
That prior work inclusing standardising columns and filtering data of diseases we aim to include, and actual cases, not suspicion
```{r}
# Data organized from raw health datasets for each country:
load(here("data", "processed", "health_data.Rdata"))
```

Shapefile of selected amazon districts of the amazon:
```{r}
our_amazon <- st_read(here("data", "raw", "amazon_districts_ALBERS.shp")) 
```

# Join data

## HEALTH
```{r}
combined_data <- our_amazon %>% as.data.frame() %>%
  distinct(COD, .keep_all = TRUE) %>%  # Keep only unique rows on COD (e.g. SR)
  dplyr::select(COD) %>% arrange(COD) %>%
  left_join(master_COD, by= "COD")

## next remove absent health data, as expected not all municipalities had health entries
# use unmatched country column, only filled on health data, not amazon districts shp
 combined_data <- combined_data %>%
  drop_na(country)
```

## Population

```{r}
pop_data <- read.csv(here("data", "raw", "ourAmz_pop_00_20.csv"))

combined_data <- combined_data %>%
  left_join(pop_data, by= c( "COD", "year"))
```

## Landscape cover and fragmentation data

Data extracted on ArcGIS and fragstats on forest cover and configuration for each municipality 

```{r}
frag_data <- read.csv(here("data", "raw", "frag_all.csv")) 

combined_data <- combined_data %>%
  left_join(frag_data, by= c("COD", "year")) 
```

## Fire incidence

```{r}
fire_data <- read.csv(here("data", "raw", "fire_all.csv"))

combined_data <- combined_data %>%
  left_join(fire_data, by= c("COD", "year"))
```

# Fire pollutants (PM2.5)

```{r}
pm25_data <- read.csv(here("data", "raw", "pm25_all.csv"))

combined_data <- combined_data %>%
  left_join(pm25_data, by= c("COD", "year"))
```

# Data structure and definitions

```{r}
combined_data <- combined_data %>%
  dplyr::select(
    # Sampling/organizing variables and population data
    country, COD, year, pop,
    # Main groups of diseases and their respective numbers
    respiratory,
    cardiovascular,
    zoonotic,
    # then individual zoonotic diseases
    chagas, hantavirus, cutaneous_leishmaniasis, visceral_leishmaniasis, rickettsial_disease, malaria,
    # Environmental data
    ## Land use and forest fragmentation
    for_PLAND, for_PD, for_ED, for_AI, 
    savanna_PLAND, notforest_PLAND,
    ## fire events and fire pollutants
    fire_MIN, fire_MAX, fire_MEAN, fire_COUNT, fire_SUM, pm25_SUM
  )

# save final raw data:
write.csv(combined_data,here("data", "processed", "amz_health_env.csv"), row.names = FALSE)
```

Now organize metadata, a data frame for the data structure and definitions

```{r}
data_structure <- tibble::tibble(
  Variable = c(
    # Administrative Metadata
    "country", "COD", "year", "pop",

    # Disease groups
    "respiratory", "cardiovascular", "zoonotic",

    # Individual zoonotic diseases
    "chagas", "hantavirus", "cutaneous_leishmaniasis", 
    "visceral_leishmaniasis", "rickettsial_disease", "malaria",

    # Landscape Cover and Fragmentation
    "for_PLAND", "savanna_PLAND", "notforest_PLAND", 
    "for_PD", "for_ED", "for_AI",

    # Fire Events
    "fire_MIN", "fire_MAX", "fire_MEAN", "fire_COUNT", "fire_SUM",

    # Air Pollution
    "pm25_SUM"
  ),

  Definition = c(
    "Country of the observation (e.g., Brazil, Peru)",
    "Geographic identifier of the administrative unit (text)",
    "Year of observation (YYYY)",
    "Estimated population of the municipality (number of people)",

    "Total reported respiratory disease cases (cases per year)",
    "Total reported cardiovascular disease cases (cases per year)",
    "Total reported zoonotic and vector-borne disease cases (cases per year)",

    "Reported cases of Chagas disease (cases per year)",
    "Reported cases of hantavirus infection (cases per year)",
    "Reported cases of cutaneous leishmaniasis (cases per year)",
    "Reported cases of visceral leishmaniasis (cases per year)",
    "Reported cases of rickettsial diseases (cases per year)",
    "Reported cases of malaria (cases per year)",

    "Natural Forest cover in municipality (PLAND metric, %)",
    "Savanna formation cover in municipality (PLAND metric, %)",
    "Non-forest natural formations (PLAND metric, %)",
    "Density of forest patches (patches per 100 ha)",
    "Forest edge density (m/ha)",
    "Forest Aggregation Index (0–100, unitless)",

    "Minimum annual fire events (pixels with fire)",
    "Maximum annual fire events (pixels/year)",
    "Mean annual fire events (pixels/year)",
    "Years with recorded fire (count)",
    "Cumulative fire events over period (sum of pixels)",

    "Sum of annual average PM2.5 concentration (μg/m³)"
  )
)


# Render the table using kable and kableExtra for a polished look
kable(data_structure, format = "html", escape = FALSE) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```


```{r, echo= FALSE}
write.csv(data_structure, here("outputs", "tables", "metadata.csv"), 
          row.names = FALSE)
```

