---
title: "Countries health data exploration and plots for Amazon project"
author: "Julia Barreto"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    thumbnails: false
    lightbox: true
    gallery: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = F, message = F, warning = F, error= F)
library(here)
library(skimr)
library(tidyverse)
library(cowplot)
theme_set(theme_cowplot())
library(kableExtra)
library(readxl)
library(ggplot2)
library(patchwork)
library(sf)
library(mapview)
# library(janitor)
```


# Loadind and overviewing data

Wrangled and organized health data.

```{r}
# load(here("data", "R01", "tidy_health.Rdata"))
combined_data <- read.csv(here("data", "combined_data.csv"))

# Structure of the data
str(combined_data)
```


```{r}
skim(combined_data)
```

Summarize by country and disease:
```{r}
country_disease_summary <- combined_data %>%
  pivot_longer(
    cols = c(respiratory, cardiovascular, chagas:malaria), 
    names_to = "disease", 
    values_to = "cases"
  ) %>%
  group_by(country, disease) %>%
  summarise(
    total_cases = if (all(is.na(cases))) NA else sum(cases, na.rm = TRUE),
    avg_cases_per_year = if (all(is.na(cases))) NA else mean(cases, na.rm = TRUE),
    max_cases = if (all(is.na(cases))) NA else max(cases, na.rm = TRUE),
    year_max_cases = if (all(is.na(cases))) NA else year[which.max(cases)],  # Year with the highest cases
    year_min_cases = if (all(is.na(cases))) NA else year[which.min(cases)],  # Year with the lowest cases
    year_range = if (all(is.na(cases))) NA else paste0(
      min(year[!is.na(cases)], na.rm = TRUE), "-", 
      max(year[!is.na(cases)], na.rm = TRUE)
    ),  # Time range of non-NA data
    .groups = "drop"
  )

# View the summary
print(country_disease_summary)
```
Summarize by disease across all countries and years:

```{r}
disease_summary <- combined_data %>%
  pivot_longer(
    cols = c(respiratory, cardiovascular, chagas:malaria), 
    names_to = "disease", 
    values_to = "cases"
  ) %>%
  group_by(disease) %>%
  summarise(
    total_cases = sum(cases, na.rm = TRUE),
    avg_cases_per_year = if (all(is.na(cases))) NA else mean(cases, na.rm = TRUE),
    max_cases = if (all(is.na(cases))) NA else max(cases, na.rm = TRUE),
    num_countries = n_distinct(country[!is.na(cases)]),  # Number of countries with data
    year_range = paste0(
      min(year[!is.na(cases)], na.rm = TRUE), "-", 
      max(year[!is.na(cases)], na.rm = TRUE)
    ),  # Year range of non-NA data
    .groups = "drop"
  )

# View the summary
print(disease_summary)
```

# HEALTH

```{r}
country_disease_year <- combined_data %>%
  ungroup() %>%
  group_by(country, year) %>%
  summarise(across(c(respiratory, cardiovascular, chagas:malaria)), na.rm= TRUE, 
            .groups = "drop") %>%
  pivot_longer(c(respiratory, cardiovascular, chagas:malaria), 
               names_to = 'diseases', 
               values_to = 'n_cases') %>%
  dplyr::select(country, year, diseases, n_cases)

complete_data <- country_disease_year %>%
  dplyr::ungroup() %>%
  tidyr::complete(
    country,
    diseases,
    year = 1995:2023,
    fill = list(n_cases = NA)
  ) %>%
  mutate(
    diseases = factor(diseases, levels = c(
      "respiratory",
      "cardiovascular",
      "chagas",
      "hantavirus",
      "cutaneous_leishmaniasis",
      "visceral_leishmaniasis",
      "rickettsia",
      "malaria")),
    country = factor(country, levels = rev(sort(unique(country))))) %>% 
  drop_na(diseases)


# Define new labels for diseases
disease_labels <- c(
  "respiratory" = "Respiratory",
  "cardiovascular" = "Cardiovascular",
  "chagas" = "Chagas",
  "hantavirus" = "Hantavirus",
  "cutaneous_leishmaniasis" = "Cutaneous Leishmaniasis",
  "visceral_leishmaniasis" = "Visceral Leishmaniasis",
  "rickettsia" = "Rickettsia",
  "malaria" = "Malaria"
)

# Define the color palette for diseases
disease_colors <- c(
  "respiratory" = "#990000",
  "cardiovascular" = "#FF9933",
  "chagas" = "#A9DFBF",
  "hantavirus" = "#58D68D",
  "cutaneous_leishmaniasis" = "#28B463",
  "visceral_leishmaniasis" = "#1D8348",
  "rickettsia" = "#117A65",
  "malaria" = "#0E6655"
)

ggplot(complete_data, aes(x = year, y = country, color = diseases)) +
  geom_line(data = subset(complete_data, !is.na(n_cases))) +
  geom_point(data = subset(complete_data, !is.na(n_cases))) + # Display only where data exists
  facet_wrap(~ diseases, labeller = labeller(diseases = disease_labels), ncol = 3) +
  scale_color_manual(values = disease_colors) +
  labs(x = "Year",
    y = "Country") +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none" # Remove the legend box
  )


ggsave(here("outputs", "diseases_countries_years.png"), width = 9, height = 7)
```




```{r, fig.height= 5, fig.width= 6}
# Aggregate the data at the country level
country_disease_data <- combined_data %>%
  group_by(country) %>%
  summarise(
    total_zoo = unique(n_zoo),
    total_resp = unique(n_resp),
    total_card = unique(n_card)
  ) %>%
  pivot_longer(cols = starts_with("total"), 
               names_to = "disease_group", 
               values_to = "total_diseases")

# Customize disease group names for the plot
country_disease_data$disease_group <- factor(country_disease_data$disease_group,
                                             levels = c("total_resp", "total_card", "total_zoo"),
                                             labels = c("Respiratory", "Cardiovascular", "Zoonotic/V-B"))

country_disease_data$country <- factor(country_disease_data$country, 
                                       levels = rev(sort(unique(country_disease_data$country))))



# Create the bar plot
country_plot <- ggplot(country_disease_data, 
                       aes(x = country, y = total_diseases, fill = disease_group)) +
  geom_bar(stat = "identity") +
  labs(y = "Total # of Cases", x = "Country", fill = "Disease Group") +
  scale_fill_manual(values = c('#990000', '#FF9933', '#006633')) +
  geom_text(aes(label = total_diseases), 
            position = position_stack(vjust = 0.5),  # Adjusting the position
            size = 4, 
            color = "white") +  # Adding text labels with the counts
  coord_flip() +
  labs(x = " ",
       y = "Number of diseases",
       fill = " ") + 
  theme(legend.position = c(0.86, 0.03),  # Adjusting legend position
        legend.justification = "bottom",  # Adjusting justification
        legend.box.just = "right")


# Display the plot
print(country_plot)

```

```{r}

# Adjust data to calculate proportions correctly
zoonotic_data_pie <- combined_data %>%
  select(country, chagas, hantavirus, cutaneous_leishmaniasis, visceral_leishmaniasis, rickettsia, malaria) %>%
  group_by(country) %>%
  summarise(across(everything(), ~ sum(.x, na.rm = TRUE))) %>%
  pivot_longer(cols = -country, 
               names_to = "disease", 
               values_to = "total_diseases") %>%
  group_by(country) %>%  # Ensure proportions are calculated per country
  mutate(
    total_diseases = ifelse(is.na(total_diseases), 0, total_diseases),  # Handle NA values
    proportion = total_diseases / sum(total_diseases) * 100,        # Calculate proportions for each country
    disease = factor(disease, 
                     levels = unique(disease),
                     labels = c("Chagas", "Hantavirus", "Cutaneous Leish.", 
                                "Visceral Leish.", "Rickettsia", "Malaria"))
  ) %>%
  ungroup()  # Ungroup after calculations
```

```{r}
# Create proportional pie charts with vertical arrangement
pie_chart_vertical <- ggplot(zoonotic_data_pie, aes(x = "", y = proportion, fill = disease)) +
  geom_bar(stat = "identity", width = 1.5) +  # Increased width for larger pies
  coord_polar(theta = "y") +
  facet_grid(country ~ ., switch = "y") +  # Align pies vertically with strip text on the left
  scale_fill_manual(values = c('#A9DFBF', '#58D68D', '#28B463', '#1D8348', '#117A65', '#0E6655')) +
  labs(fill = "", title = "Zoonotic/Vector-Borne Diseases") +
  theme_void() +  # Simplify chart for clean pie visualization
  theme(
    legend.position = "right",  # Place the legend to the right
    legend.text = element_text(size = 10),
    strip.text.y = element_text(size = 12, face = "bold", angle = 0),  # Larger strip text for countries
    plot.margin = margin(10, 10, 10, 10)  # Add space around the plot
  )

# Display the updated pie chart
print(pie_chart_vertical)

```

```{r, fig.width= 10}
country_plot + pie_chart_vertical + plot_layout(widths = c(3,1))
ggsave(here("outputs", "diseases_countries.png"), width = 11, height = 5)
```

## Number of cases per country
```{r, fig.width= 30, fig.height= 20}
# Define the full range of years
full_years <- seq(2000, max(combined_data$year))
 
grouped_data <- combined_data %>%
  select(country, year, respiratory, cardiovascular, zoonotic) %>%
  group_by(country, year) %>%
  summarise(
    total_resp = sum(respiratory, na.rm = TRUE),
    total_card = sum(cardiovascular, na.rm = TRUE),
    total_zoo = sum(zoonotic, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = starts_with("total"), 
               names_to = "disease_group", 
               values_to = "total_cases") %>%
  mutate(
    disease_group = factor(disease_group,
                           levels = c("total_resp", "total_card", "total_zoo"),
                           labels = c("Respiratory", "Cardiovascular", "Zoonotic/V-B"))
  )

country_plots <- grouped_data %>%
  group_split(country) %>%
  setNames(unique(grouped_data$country)) %>%
  lapply(function(data) {
    ggplot(data, aes(x = year, y = total_cases, fill = disease_group)) +
      geom_bar(stat = "identity", position = "dodge", width = 0.8) +  # Adjust bar width here
      scale_fill_manual(values = c('#990000', '#FF9933', '#006633')) +
      scale_x_continuous(
        breaks = seq(min(data$year), max(data$year), by = 5),  # Show ticks every 5 years
        labels = seq(min(data$year), max(data$year), by = 5)
      ) +
      labs(title = unique(data$country)) +  # Add country name as the title
      theme(
        legend.title = element_blank(),
        legend.position = "top",  # Place legends above plots
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 12, face = "bold", hjust = 0.5)  # Center country titles
      ) + 
      labs(y = "Total # of Cases", x = "Year")
  })

# Save plots to files or display them
for (i in seq_along(country_plots)) {
  ggsave(
    filename = paste0("outputs/cases_per_year_", unique(grouped_data$country)[i], ".png"),
    plot = country_plots[[i]],
    width = 8, height = 6
  )
}

# Combine all plots into a single panel
panel <- wrap_plots(country_plots, ncol = 3) +
  plot_layout(guides='collect') + plot_annotation(tag_levels = "a")  &
  theme(legend.position='right') &
  labs(x = NULL, y = NULL) # Suppress individual plot titles

  

add_global_label <- function(pwobj, Xlab = NULL, Ylab = NULL, Xgap = 0.03, Ygap = 0.03, ...) {
  ylabgrob <- patchwork::plot_spacer()
  if (!is.null(Ylab)) {
    ylabgrob <- ggplot() +
      geom_text(aes(x = .5, y = .5), label = Ylab, angle = 90, ...) +
      theme_void()
  }
  if (!is.null(Xlab)) {
    xlabgrob <- ggplot() +
      geom_text(aes(x = .5, y = .5), label = Xlab, ...) +
      theme_void()
  }
  if (!is.null(Ylab) & is.null(Xlab)) {
    return((ylabgrob + patchworkGrob(pwobj)) + 
             patchwork::plot_layout(widths = 100 * c(Ygap, 1 - Ygap)))
  }
  if (is.null(Ylab) & !is.null(Xlab)) {
    return((ylabgrob + pwobj) + 
             (xlabgrob) +
             patchwork::plot_layout(heights = 100 * c(1 - Xgap, Xgap),
                                    widths = c(0, 100),
                                    design = "
                                   AB
                                   CC
                                   "
             ))
  }
  if (!is.null(Ylab) & !is.null(Xlab)) {
    return((ylabgrob + pwobj) + 
             (xlabgrob) +
             patchwork::plot_layout(heights = 100 * c(1 - Xgap, Xgap),
                                    widths = 100 * c(Ygap, 1 - Ygap),
                                    design = "
                                   AB
                                   CC
                                   "
             ))
  }
  return(pwobj)
}


final_panel <- panel %>%
  add_global_label(Ylab = "Total # of Cases",
                   Xlab = "Year",
                         size = 5,
                        Ygap = 0.04) + 
  plot_annotation(tag_levels = 'a')

print(final_panel)

ggsave(here("outputs", "cases_countries.png"), width = 11, height = 9)
```

## Individual zoonotic and vector-borne diseases

```{r, fig.width = 10, fig.height = 8}
# Prepare data for zoonotic diseases with a consistent x-axis range
zoonotic_data <- combined_data %>%
  select(country, year, chagas, hantavirus, cutaneous_leishmaniasis, visceral_leishmaniasis, rickettsia, malaria) %>%
  pivot_longer(cols = -c(country, year), 
               names_to = "disease", 
               values_to = "total_cases") %>%
  complete(country, year = full_years, disease, fill = list(total_cases = 0)) %>%  # Fill missing years with 0
  group_by(country, year, disease) %>%
  summarise(total_cases = sum(total_cases, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    disease = factor(disease,
                     levels = c("chagas", "hantavirus", "cutaneous_leishmaniasis", 
                                "visceral_leishmaniasis", "rickettsia", "malaria"),
                     labels = c("Chagas", "Hantavirus", "Cutaneous Leishmaniasis", 
                                "Visceral Leishmaniasis", "Rickettsia", "Malaria"))
  )

# Define the color palette
zoonotic_palette <- c('#A9DFBF', '#58D68D', '#28B463', '#1D8348', '#117A65', '#0E6655')

# Create individual plots for each country
zoonotic_plots <- zoonotic_data %>%
  group_split(country) %>%
  setNames(unique(zoonotic_data$country)) %>%
  lapply(function(data) {
    ggplot(data, aes(x = factor(year), y = total_cases, fill = disease)) +
      geom_bar(stat = "identity", position = "dodge", width = 1) +  # Consistent bar width
      scale_fill_manual(values = zoonotic_palette) +
      scale_x_discrete(
        breaks = as.character(seq(2000, max(full_years), by = 5)),  # Show ticks every 5 years
        labels = seq(2000, max(full_years), by = 5)
      ) +
      labs(title = unique(data$country)) +
      theme(
        legend.title = element_blank(),
        legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
      ) + 
      labs(y = "Total # of Cases", x = "Year")
  })

# Combine all plots into a single panel
zoonotic_panel <- wrap_plots(zoonotic_plots, ncol = 3) +
  plot_layout(guides = "collect") + plot_annotation(tag_levels = "a")  & 
  theme(legend.position = "right") &
  labs(x = NULL, y = NULL)


# Add shared y-axis label using `add_global_label`
final_zoonotic_panel <- zoonotic_panel %>%
  add_global_label(
    Ylab = "Total # of Cases",
    Xlab = "Year",
    size = 5,
    Ygap = 0.04
  ) + 
  plot_annotation(tag_levels = 'a')

# Display the final panel
print(final_zoonotic_panel)

# Save the combined panel
ggsave(here("outputs", "zoonotic_cases_countries.png"), width = 10, height = 8)
```

# ENVIRONMENTAL INDICATORS

Summary of environmental indicators and data sources
```{r}
data_structure <- tibble::tibble(
  Indicator = c(
    "Forest Cover Metrics", 
    "Fragmentation Metrics", 
    "Savanna Cover Metrics", 
    "Fire-related Indicators", 
    "PM2.5 (Calibrated)", 
    "Burned Area", 
    "Indigenous Territories"
  ),
  Description = c(
    "Proportion of forest cover within and outside Indigenous Territories (ITs)", 
    "Forest edge density, patch density, and aggregation index", 
    "Proportion of savanna cover within municipalities", 
    "Number of fire events and related pollution (PM2.5)", 
    "PM2.5 concentrations derived from MAIAC AOD and calibrated with SEDAC data", 
    "Accumulated burned area (sum of pixels burned at least once) and burn frequency", 
    "Status (recognized/unrecognized), boundaries, and coverage of ITs"
  ),
  Source = c(
    "MapBiomas Collection 4", 
    "MapBiomas Collection 4, Fragstats", 
    "MapBiomas Collection 4", 
    "MODIS Terra (MOD14A1), MAIAC AOD, NASA SEDAC", 
    "MAIAC AOD, NASA SEDAC", 
    "MODIS Terra (MOD14A1)", 
    "RAISG"
  ),
  Spatial_Resolution = c(
    "30 m", 
    "30 m", 
    "30 m", 
    "1 km", 
    "1 km", 
    "1 km", 
    "IT boundaries"
  )
)

# Render the table using kable and kableExtra for a polished look
data_structure %>%
  kable("html", escape = FALSE, col.names = c("Indicator", "Description", "Source", "Spatial Resolution")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# Export the table to a CSV file
write.csv(data_structure, here("data", "summary_indicators.csv"), row.names = FALSE)
```


## Plot showing temporal change landscape
(forest cover, fragmentation, and fire pollutants.

```{r}
# color gradient for each country, from darkest to lightest
color_palette <- c("#3E2723", "#5D4037", "#795548", "#8D6E63", "#A1887F", "#BCAAA4", "#D7CCC8")

country_order <- c("Brasil", "Peru", "Bolivia", "Colombia", "Venezuela", "Suriname", "Ecuador")
color_mapping <- data.frame(country = country_order, color_code = color_palette)
```


### Forest Cover (for_PLAND)
```{r}
data_with_colors <- combined_data %>%
  dplyr::left_join(color_mapping, by = "country") %>%
  group_by(country, year) %>%
  summarise(
    for_PLAND_mean = mean(for_PLAND, na.rm = TRUE),
    for_PLAND_sd = sd(for_PLAND, na.rm = TRUE),
    color_code = unique(color_code)  # Retain color code after grouping
  ) %>%
  ungroup() %>%
  mutate(
    country = case_when(
      country == "Bolivia" ~ "BO",
      country == "Brasil" ~ "BR",
      country == "Colombia" ~ "CO",
      country == "Ecuador" ~ "EC",
      country == "Guiana FR" ~ "GUF",
      country == "Peru" ~ "PE",
      country == "Suriname" ~ "SR",
      country == "Venezuela" ~ "VE",
      TRUE ~ country  # Keep original value for unmatched cases
    ),
    for_PLAND_mean = ifelse(is.nan(for_PLAND_mean), NA, for_PLAND_mean),
    for_PLAND_sd = ifelse(is.nan(for_PLAND_sd), NA, for_PLAND_sd)
  ) %>% 
  drop_na(for_PLAND_mean)

# Find the last year for each country to place the labels
label_data <- data_with_colors %>%
  group_by(country) %>%
  filter(year == max(year)) %>%
  ungroup()

# Plotting with ribbons for standard deviation and country labels
(plot_forest_cover <- ggplot(data_with_colors, aes(x = year, y = for_PLAND_mean, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
  # geom_ribbon(
  #   aes(
  #     ymin = pmax(for_PLAND_mean - for_PLAND_sd, 0),  # Ensure ymin is at least 0
  #     ymax = pmin(for_PLAND_mean + for_PLAND_sd, 100),  # Ensure ymax does not exceed 100
  #     fill = color_code
  #   ),
  #   alpha = 0.1, color = NA
  # ) +
  scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
  breaks = seq(2000, 2020, by = 1),  # Ensure all years have a tick
  labels = seq(2000, 2020, by = 1),  # Explicitly define labels for all ticks
  expand = c(0, 0)  # No padding on the x-axis
) +
  scale_y_continuous(
    limits = c(0, 100),  # Fix y-axis range from 0 to 100
    expand = c(0, 0)  # No padding on the y-axis
  ) +
  labs(
    # title = "Forest Percentage Variation Over Years by Country",
    x = "Year",
    y = "Forest Cover Percent"
  ) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),  # Increase x-axis label size
    axis.text.y = element_text(size = 16),  # Increase y-axis label size
    axis.title.x = element_text(size = 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size = 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size = 16),  # Increase legend text size
    legend.title = element_text(size = 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = for_PLAND_mean, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size = 5
  ) +
  xlim(2000, 2020))
```

### Patch Density (for_PD)
```{r}
data_with_colors <- combined_data %>%
  left_join(color_mapping, by = "country") %>%
  group_by(country, year) %>%
  summarise(
    for_PD_mean = mean(for_PD, na.rm = TRUE),
    for_PD_sd = sd(for_PD, na.rm = TRUE),
    color_code = unique(color_code)
  ) %>%
  ungroup() %>%
  mutate(
    country = case_when(
      country == "Bolivia" ~ "BO",
      country == "Brasil" ~ "BR",
      country == "Colombia" ~ "CO",
      country == "Ecuador" ~ "EC",
      country == "Guiana FR" ~ "GUF",
      country == "Peru" ~ "PE",
      country == "Suriname" ~ "SR",
      country == "Venezuela" ~ "VE",
      TRUE ~ country  # Keep original value for unmatched cases
    ),
    for_PD_mean = ifelse(is.nan(for_PD_mean), NA, for_PD_mean),
    for_PD_sd = ifelse(is.nan(for_PD_sd), NA, for_PD_sd)
  ) %>% 
  drop_na(for_PD_mean)

label_data <- data_with_colors %>%
  group_by(country) %>%
  filter(year == max(year)) %>%
  ungroup()


(plot_patch_density <- ggplot(data_with_colors, aes(x = year, y = for_PD_mean, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
  # geom_ribbon(
  #   aes(
  #     ymin = pmax(for_PD_mean - for_PD_sd, 0),  # Ensure ymin is at least 0
  #     ymax = for_PD_mean + for_PD_sd,  # No upper limit for patch density
  #     fill = color_code
  #   ),
  #   alpha = 0.1, color = NA
  # ) +
  scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
    breaks = seq(2000, 2020, by = 1),
    labels = seq(2000, 2020, by = 1),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    limits = c(0, NA),  # Start from 0, no upper limit
    expand = c(0, 0)
  ) +
  labs(
    # title = "Patch Density Over Years by Country",
    x = "Year",
    y = "Patch Density"
  ) +
  theme_cowplot() +
  
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size= 16),  # Increase x-axis label size
    axis.text.y = element_text(size= 16),  # Increase y-axis label size
    axis.title.x = element_text(size= 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size= 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size= 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size= 16),  # Increase legend text size
    legend.title = element_text(size= 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = for_PD_mean, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size= 5
  ) +
  xlim(2000, 2020))
```
### Edge Density (for_ED)
```{r}
data_with_colors <- combined_data %>%
  left_join(color_mapping, by = "country") %>%
  group_by(country, year) %>%
  summarise(
    for_ED_mean = mean(for_ED, na.rm = TRUE),
    for_ED_sd = sd(for_ED, na.rm = TRUE),
    color_code = unique(color_code)
  ) %>%
  ungroup() %>%
  mutate(
    country = case_when(
      country == "Bolivia" ~ "BO",
      country == "Brasil" ~ "BR",
      country == "Colombia" ~ "CO",
      country == "Ecuador" ~ "EC",
      country == "Guiana FR" ~ "GUF",
      country == "Peru" ~ "PE",
      country == "Suriname" ~ "SR",
      country == "Venezuela" ~ "VE",
      TRUE ~ country  # Keep original value for unmatched cases
    ),
    for_ED_mean = ifelse(is.nan(for_ED_mean), NA, for_ED_mean),
    for_ED_sd = ifelse(is.nan(for_ED_sd), NA, for_ED_sd)
  ) %>% 
  drop_na(for_ED_mean)

label_data <- data_with_colors %>%
  group_by(country) %>%
  filter(year == max(year)) %>%
  ungroup()

(plot_edge_density <- ggplot(data_with_colors, aes(x = year, y = for_ED_mean, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
  # geom_ribbon(
  #   aes(
  #     ymin = pmax(for_ED_mean - for_ED_sd, 0),  # Ensure ymin is at least 0
  #     ymax = for_ED_mean + for_ED_sd,  # No upper limit for edge density
  #     fill = color_code
  #   ),
  #   alpha = 0.1, color = NA
  # ) +
  scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
    breaks = seq(2000, 2020, by = 1),
    labels = seq(2000, 2020, by = 1),
    expand = c(0, 0)
  ) +
  # scale_y_continuous(
  #   limits = c(0, NA),  # Start from 0, no upper limit
  #   expand = c(0, 0)
  # ) +
  labs(
    # title = "Edge Density Over Years by Country",
    x = "Year",
    y = "Edge Density"
  ) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size= 16),  # Increase x-axis label size
    axis.text.y = element_text(size= 16),  # Increase y-axis label size
    axis.title.x = element_text(size= 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size= 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size= 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size= 16),  # Increase legend text size
    legend.title = element_text(size= 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = for_ED_mean, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size= 5
  ) +
  xlim(2000, 2020))
```

### Aggregation Index (for_AI)
```{r}
data_with_colors <- combined_data %>%
  left_join(color_mapping, by = "country") %>%
  group_by(country, year) %>%
  summarise(
    for_AI_mean = mean(for_AI, na.rm = TRUE),
    for_AI_sd = sd(for_AI, na.rm = TRUE),
    color_code = unique(color_code)
  ) %>%
  ungroup() %>%
  mutate(
    country = case_when(
      country == "Bolivia" ~ "BO",
      country == "Brasil" ~ "BR",
      country == "Colombia" ~ "CO",
      country == "Ecuador" ~ "EC",
      country == "Guiana FR" ~ "GUF",
      country == "Peru" ~ "PE",
      country == "Suriname" ~ "SR",
      country == "Venezuela" ~ "VE",
      TRUE ~ country  # Keep original value for unmatched cases
    ),
    for_AI_mean = ifelse(is.nan(for_AI_mean), NA, for_AI_mean),
    for_AI_sd = ifelse(is.nan(for_AI_sd), NA, for_AI_sd)
  ) %>% 
  drop_na(for_AI_mean)

label_data <- data_with_colors %>%
  group_by(country) %>%
  filter(year == max(year)) %>%
  ungroup()

(plot_aggregation <- ggplot(data_with_colors, aes(x = year, y = for_AI_mean, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
  # geom_ribbon(
  #   aes(
  #     ymin = pmax(for_AI_mean - for_AI_sd, 0),  # Ensure ymin is at least 0
  #     ymax = for_AI_mean + for_AI_sd,  # No upper limit for aggregation index
  #     fill = color_code
  #   ),
  #   alpha = 0.1, color = NA
  # ) +
  scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
    breaks = seq(2000, 2020, by = 1),
    labels = seq(2000, 2020, by = 1),
    expand = c(0, 0)
  ) +
  # scale_y_continuous(
  #   limits = c(0, NA),  # Start from 0, no upper limit
  #   expand = c(0, 0)
  # ) +
  labs(
    # title = "Aggregation Index Over Years by Country",
    x = "Year",
    y = "Aggregation Index"
  ) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size= 16),  # Increase x-axis label size
    axis.text.y = element_text(size= 16),  # Increase y-axis label size
    axis.title.x = element_text(size= 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size= 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size= 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size= 16),  # Increase legend text size
    legend.title = element_text(size= 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = for_AI_mean, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size= 5
  ) +
  xlim(2000, 2020))
```

### Fire Pollutant  (pm25_SUM)

Be attentive to Suriname, in which PM values are very high, also the only at country level or might be due to winds in the region - check.

```{r}
data_with_colors <- combined_data %>%
  left_join(color_mapping, by = "country") %>%
  group_by(country, year) %>%
  summarise(
    pm25_SUM_mean = mean(pm25_SUM, na.rm = TRUE),
    pm25_SUM_sd = sd(pm25_SUM, na.rm = TRUE),
    color_code = unique(color_code)
  ) %>%
  ungroup() %>%
  mutate(
    country = case_when(
      country == "Bolivia" ~ "BO",
      country == "Brasil" ~ "BR",
      country == "Colombia" ~ "CO",
      country == "Ecuador" ~ "EC",
      country == "Guiana FR" ~ "GUF",
      country == "Peru" ~ "PE",
      country == "Suriname" ~ "SR",
      country == "Venezuela" ~ "VE",
      TRUE ~ country  # Keep original value for unmatched cases
    ),
    pm25_SUM_mean = ifelse(is.nan(pm25_SUM_mean), NA, pm25_SUM_mean),
    pm25_SUM_sd = ifelse(is.nan(pm25_SUM_sd), NA, pm25_SUM_sd)
  ) %>% 
  drop_na(pm25_SUM_mean)

label_data <- data_with_colors %>%
  group_by(country) %>%
  filter(year == max(year)) %>%
  ungroup()

(plot_pm25 <- ggplot(data_with_colors, aes(x = year, y = pm25_SUM_mean, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
  # geom_ribbon(
  #   aes(
  #     ymin = pmax(pm25_SUM_mean - pm25_SUM_sd, 0),  # Ensure ymin is at least 0
  #     ymax = pm25_SUM_mean + pm25_SUM_sd,  # No upper limit for patch density
  #     fill = color_code
  #   ),
  #   alpha = 0.1, color = NA
  # ) +
  scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
    breaks = seq(2000, 2020, by = 1),
    labels = seq(2000, 2020, by = 1),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    # limits = c(0, 100000000000),  # Start from 0, no upper limit
    expand = c(0, 0)
  ) +
  labs(
    # title = "Fire Pollutant (PM2.5) Over Years by Country",
    x = "Year",
    y = expression("Fire Pollutant (PM"[2.5]*")")
  ) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size= 16),  # Increase x-axis label size
    axis.text.y = element_text(size= 16),  # Increase y-axis label size
    axis.title.x = element_text(size= 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size= 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size= 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size= 16),  # Increase legend text size
    legend.title = element_text(size= 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = pm25_SUM_mean, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size= 5
  ) +
  xlim(2000, 2020))
```

```{r, fig.width = 15, fig.height = 9}
# Remove x-axis labels from individual plots
plot_forest_cover <- plot_forest_cover + theme(axis.title.x = element_blank())
plot_pm25 <- plot_pm25 + theme(axis.title.x = element_blank())
plot_patch_density <- plot_patch_density + theme(axis.title.x = element_blank())
# plot_aggregation <- plot_aggregation + theme(axis.title.x = element_blank())
plot_edge_density <- plot_edge_density + theme(axis.title.x = element_blank())

# Create the panel layout with your original positioning
panel_plot <- (plot_forest_cover | plot_pm25) /
              (plot_patch_density | plot_aggregation | plot_edge_density) +
    plot_annotation(tag_levels = 'a')


panel_plot

ggsave(here("outputs", "landscape_change.png"), width = 15, height = 9)
```



