---
title: "Data exploration and display"
author: "Julia Barreto"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    thumbnails: false
    lightbox: true
    gallery: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = F, message = F, warning = F, error= F)
library(here)
library(skimr)
library(tidyverse)
library(cowplot)
theme_set(theme_cowplot())
library(kableExtra)
library(readxl)
library(ggplot2)
library(patchwork)
library(sf)
library(mapview)
# library(janitor)
```

[![DOI](https://zenodo.org/badge/1005732035.svg)](https://doi.org/10.5281/zenodo.15760516)

# Loading and overviewing data

Wrangled and organized health data.

```{r}
combined_data <- read.csv(here("data", "processed", "amz_health_env.csv"))

str(combined_data)
```


```{r}
skim(combined_data)
```
## Cross-countries yearly reported cases per disease

Summarize by country and disease:
```{r}
country_disease_summary <- combined_data %>%
  pivot_longer(
    cols = c(respiratory, cardiovascular, chagas:malaria), 
    names_to = "disease", 
    values_to = "cases"
  ) %>%
  group_by(country, disease) %>%
  summarise(
    total_cases = if (all(is.na(cases))) NA else sum(cases, na.rm = TRUE),
    avg_cases_per_year = if (all(is.na(cases))) NA else mean(cases, na.rm = TRUE),
    max_cases = if (all(is.na(cases))) NA else max(cases, na.rm = TRUE),
    year_max_cases = if (all(is.na(cases))) NA else year[which.max(cases)],  # Year with the highest cases
    year_min_cases = if (all(is.na(cases))) NA else year[which.min(cases)],  # Year with the lowest cases
    year_range = if (all(is.na(cases))) NA else paste0(
      min(year[!is.na(cases)], na.rm = TRUE), "-", 
      max(year[!is.na(cases)], na.rm = TRUE)
    ),  # Time range of non-NA data
    .groups = "drop"
  )

kable(country_disease_summary, format = "html", escape = FALSE) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

```{r}
country_disease_year <- combined_data %>%
  ungroup() %>%
  group_by(country, year) %>%
  summarise(across(c(respiratory, cardiovascular, chagas:malaria)), na.rm= TRUE, 
            .groups = "drop") %>%
  pivot_longer(c(respiratory, cardiovascular, chagas:malaria), 
               names_to = 'diseases', 
               values_to = 'n_cases') %>%
  dplyr::select(country, year, diseases, n_cases)

complete_data <- country_disease_year %>%
  dplyr::ungroup() %>%
  tidyr::complete(
    country,
    diseases,
    year = 1995:2023,
    fill = list(n_cases = NA)
  ) %>%
  mutate(
    diseases = factor(diseases, levels = c(
      "respiratory",
      "cardiovascular",
      "chagas",
      "hantavirus",
      "cutaneous_leishmaniasis",
      "visceral_leishmaniasis",
      "rickettsial_disease",
      "malaria")),
    country = factor(country, levels = rev(sort(unique(country))))) %>% 
  drop_na(diseases)


# Define new labels for diseases
disease_labels <- c(
  "respiratory" = "Respiratory",
  "cardiovascular" = "Cardiovascular",
  "chagas" = "Chagas",
  "hantavirus" = "Hantavirus",
  "cutaneous_leishmaniasis" = "Cutaneous leishmaniasis",
  "visceral_leishmaniasis" = "Visceral leishmaniasis",
  "rickettsial_disease" = "Rickettsial diseases",
  "malaria" = "Malaria"
)

# Define the color palette for diseases
disease_colors <- c(
  "respiratory" = "#990000",
  "cardiovascular" = "#FF9933",
  "chagas" = "#A9DFBF",
  "hantavirus" = "#58D68D",
  "cutaneous_leishmaniasis" = "#28B463",
  "visceral_leishmaniasis" = "#1D8348",
  "rickettsial_disease" = "#117A65",
  "malaria" = "#0E6655"
)

ggplot(complete_data, aes(x = year, y = country, color = diseases)) +
  geom_line(data = subset(complete_data, !is.na(n_cases))) +
  geom_point(data = subset(complete_data, !is.na(n_cases))) + # Display only where data exists
  facet_wrap(~ diseases, labeller = labeller(diseases = disease_labels), ncol = 3) +
  scale_color_manual(values = disease_colors) +
  labs(x = "Year",
    y = "Country") +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none" # Remove the legend box
  )


ggsave(here("outputs", "figures", "diseases_countries_years.jpg"), width = 9, height = 7)
```

## Pan-Amazon yearly reported cases per disease
Summarize by disease across all countries and years:

```{r}
disease_summary <- combined_data %>%
  pivot_longer(
    cols = c(respiratory, cardiovascular, chagas:malaria), 
    names_to = "disease", 
    values_to = "cases"
  ) %>%
  group_by(disease) %>%
  summarise(
    total_cases = sum(cases, na.rm = TRUE),
    avg_cases_per_year = if (all(is.na(cases))) NA else mean(cases, na.rm = TRUE),
    max_cases = if (all(is.na(cases))) NA else max(cases, na.rm = TRUE),
    num_countries = n_distinct(country[!is.na(cases)]),  # Number of countries with data
    year_range = paste0(
      min(year[!is.na(cases)], na.rm = TRUE), "-", 
      max(year[!is.na(cases)], na.rm = TRUE)
    ),  # Year range of non-NA data
    .groups = "drop"
  )

# View the summary
print(disease_summary)
```

# HEALTH



```{r, fig.height= 5, fig.width= 6}
# extract number of individual diseases from control/summary data table
control0 <- read.csv(here("outputs", "tables", "health_summary.csv"))

control1 <- control0 %>%
  dplyr::select(country, disease_group, diseases) %>%
  mutate(across(everything(), ~na_if(.x, "-"))) %>% # this makes Guyana actual NA
  drop_na(disease_group) %>%
  fill(country) # necessary as i've improvised to merged country names
  
control <- control1 %>%  # Count number of diseases per group per country:
  mutate(n_diseases = str_count(diseases, ";") + 1) %>%
  group_by(country, disease_group) %>%
  summarise(n = sum(n_diseases), .groups = "drop") %>%
  pivot_wider(names_from = disease_group,
              values_from = n,
              names_prefix = "n_")

country_disease_data <- control %>%
  group_by(country) %>%
  summarise(
    total_zoo = unique(n_zoonotic),
    total_resp = unique(n_respiratory),
    total_card = unique(n_cardiovascular)
  ) %>%
  pivot_longer(cols = starts_with("total"), 
               names_to = "disease_group", 
               values_to = "total_diseases")

# Customize disease group names for the plot
country_disease_data$disease_group <- 
  factor(country_disease_data$disease_group,
         levels = c("total_resp", "total_card", "total_zoo"),
         labels = c("Respiratory", "Cardiovascular", "Zoonotic/V-B"))

country_disease_data$country <- 
  factor(country_disease_data$country, 
         levels = rev(sort(unique(country_disease_data$country))))

# Create the bar plot
country_plot <- ggplot(country_disease_data, 
                       aes(x = country, y = total_diseases, fill = disease_group)) +
  geom_bar(stat = "identity") +
  labs(y = "Total # of Cases", x = "Country", fill = "Disease Group") +
  scale_fill_manual(values = c('#990000', '#FF9933', '#006633')) +
  geom_text(aes(label = total_diseases), 
            position = position_stack(vjust = 0.5),  # Adjusting the position
            size = 4, 
            color = "white") +  # Adding text labels with the counts
  coord_flip() +
  labs(x = " ",
       y = "Number of diseases",
       fill = " ") + 
  theme(legend.position = c(0.77, 0.03),  # Adjusting legend position
        legend.justification = "bottom",  # Adjusting justification
        legend.box.just = "right")


# Display the plot
print(country_plot)
```

```{r}
# Adjust data to calculate proportions correctly
zoonotic_data_pie <- combined_data %>%
  select(country, chagas, hantavirus, cutaneous_leishmaniasis, visceral_leishmaniasis, rickettsial_disease, malaria) %>%
  group_by(country) %>%
  summarise(across(everything(), ~ sum(.x, na.rm = TRUE))) %>%
  pivot_longer(cols = -country, 
               names_to = "disease", 
               values_to = "total_diseases") %>%
  group_by(country) %>%  # Ensure proportions are calculated per country
  mutate(
    total_diseases = ifelse(is.na(total_diseases), 0, total_diseases),  # Handle NA values
    proportion = total_diseases / sum(total_diseases) * 100,        # Calculate proportions for each country
    disease = factor(disease, 
                     levels = unique(disease),
                     labels = c("Chagas", "Hantavirus", "Cutaneous leishmaniasis", 
                                "Visceral leishmaniasis", "Rickettsial diseases", "Malaria"))
  ) %>%
  ungroup()  # Ungroup after calculations

pie_chart_vertical <- ggplot(zoonotic_data_pie, aes(x = "", y = proportion, fill = disease)) +
  geom_bar(stat = "identity", width = 1.5) +  # Increased width for larger pies
  coord_polar(theta = "y") +
  facet_grid(country ~ ., switch = "y") +  # Align pies vertically with strip text on the left
  scale_fill_manual(values = c('#A9DFBF', '#58D68D', '#28B463', '#1D8348', '#117A65', '#0E6655')) +
  labs(fill = "", title = "Zoonotic/Vector-Borne Diseases") +
  theme_void() +  # Simplify chart for clean pie visualization
  theme(
    legend.position = "right",  # Place the legend to the right
    legend.text = element_text(size = 10),
    strip.text.y = element_text(size = 12, face = "bold", angle = 0),  # Larger strip text for countries
    plot.margin = margin(10, 10, 10, 10)  # Add space around the plot
  )

# Display the updated pie chart
print(pie_chart_vertical)

```

```{r, fig.width= 8}
country_plot + pie_chart_vertical + plot_layout(widths = c(3,1))
ggsave(here("outputs", "figures", "diseases_countries.jpg"), width = 10, height = 6)
```

## Number of cases per country
```{r, fig.width= 20, fig.height= 20}
# Define the full range of years
full_years <- seq(2000, max(combined_data$year))
 
grouped_data <- combined_data %>%
  select(country, year, respiratory, cardiovascular, zoonotic) %>%
  group_by(country, year) %>%
  summarise(
    total_resp = sum(respiratory, na.rm = TRUE),
    total_card = sum(cardiovascular, na.rm = TRUE),
    total_zoo = sum(zoonotic, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = starts_with("total"), 
               names_to = "disease_group", 
               values_to = "total_cases") %>%
  mutate(
    disease_group = factor(disease_group,
                           levels = c("total_resp", "total_card", "total_zoo"),
                           labels = c("Respiratory", "Cardiovascular", "Zoonotic/V-B"))
  )

country_plots <- grouped_data %>%
  group_split(country) %>%
  setNames(unique(grouped_data$country)) %>%
  lapply(function(data) {
    ggplot(data, aes(x = year, y = total_cases, fill = disease_group)) +
      geom_bar(stat = "identity", position = "dodge", width = 0.8) +  # Adjust bar width here
      scale_fill_manual(values = c('#990000', '#FF9933', '#006633')) +
      scale_x_continuous(
        breaks = seq(min(data$year), max(data$year), by = 5),  # Show ticks every 5 years
        labels = seq(min(data$year), max(data$year), by = 5)
      ) +
      labs(title = unique(data$country)) +  # Add country name as the title
      theme(
        legend.title = element_blank(),
        legend.position = "top",  # Place legends above plots
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 12, face = "bold", hjust = 0.5)  # Center country titles
      ) + 
      labs(y = "Total # of Cases", x = "Year")
  })

# Save plots to files or display them
for (i in seq_along(country_plots)) {
  ggsave(
    filename = paste0("outputs/figures/cases_per_year_", unique(grouped_data$country)[i], ".jpg"),
    plot = country_plots[[i]],
    width = 8, height = 6
  )
}

# Combine all plots into a single panel
panel <- wrap_plots(country_plots, ncol = 3) +
  plot_layout(guides='collect') + plot_annotation(tag_levels = "a")  &
  theme(legend.position='right') &
  labs(x = NULL, y = NULL) # Suppress individual plot titles

  

add_global_label <- function(pwobj, Xlab = NULL, Ylab = NULL, Xgap = 0.03, Ygap = 0.03, ...) {
  ylabgrob <- patchwork::plot_spacer()
  if (!is.null(Ylab)) {
    ylabgrob <- ggplot() +
      geom_text(aes(x = .5, y = .5), label = Ylab, angle = 90, ...) +
      theme_void()
  }
  if (!is.null(Xlab)) {
    xlabgrob <- ggplot() +
      geom_text(aes(x = .5, y = .5), label = Xlab, ...) +
      theme_void()
  }
  if (!is.null(Ylab) & is.null(Xlab)) {
    return((ylabgrob + patchworkGrob(pwobj)) + 
             patchwork::plot_layout(widths = 100 * c(Ygap, 1 - Ygap)))
  }
  if (is.null(Ylab) & !is.null(Xlab)) {
    return((ylabgrob + pwobj) + 
             (xlabgrob) +
             patchwork::plot_layout(heights = 100 * c(1 - Xgap, Xgap),
                                    widths = c(0, 100),
                                    design = "
                                   AB
                                   CC
                                   "
             ))
  }
  if (!is.null(Ylab) & !is.null(Xlab)) {
    return((ylabgrob + pwobj) + 
             (xlabgrob) +
             patchwork::plot_layout(heights = 100 * c(1 - Xgap, Xgap),
                                    widths = 100 * c(Ygap, 1 - Ygap),
                                    design = "
                                   AB
                                   CC
                                   "
             ))
  }
  return(pwobj)
}


final_panel <- panel %>%
  add_global_label(Ylab = "Total # of Cases",
                   Xlab = "Year",
                         size = 5,
                        Ygap = 0.04) + 
  plot_annotation(tag_levels = 'a')

print(final_panel)

ggsave(here("outputs", "figures", "cases_countries.jpg"), width = 11, height = 9)
```

# ENVIRONMENTAL INDICATORS

Summary of environmental indicators and data sources
```{r}
data_structure <- tibble::tibble(
  Indicator = c(
    "Forest Cover Metric", 
    "Fragmentation Metrics", 
    "Savanna Cover Metric", 
    "Non-forest Cover Metric", 
    "Fire Events Indicators", 
    "PM2.5 (Calibrated)"
  ),
  Description = c(
    "Proportion of natural forest cover within and outside municipalities ('for_PLAND')",
    "Forest edge density ('for_ED'), patch density ('for_PD'), and aggregation index ('for_AI')",
    "Proportion of savanna formation cover within municipalities ('savanna_PLAND')",
    "Proportion of non-forest natural formation cover within municipalities ('notforest_PLAND')",
    "Indicators of numbers of fire events ('fire_MIN', 'fire_MAX', 'fire_MEAN'), and accumulated burned area (sum of pixels burned at least once; 'fire_SUM') and burn frequency ('fire_COUNT')",
    "PM2.5 concentrations derived from MAIAC AOD and calibrated with SEDAC data ('pm25_SUM')"
  ),
  Source = c(
    "MapBiomas Collection 4 (code 1, 3 and 6)", 
    "MapBiomas Collection 4, Fragstats", 
    "MapBiomas Collection 4 (code 4)",
    "MapBiomas Collection 4 (code 10)",
    "MODIS Terra (MOD14A1), MAIAC AOD, NASA SEDAC", 
    "MAIAC AOD, NASA SEDAC"
  ),
  Spatial_Resolution = c(
    "30 m", 
    "30 m", 
    "30 m", 
    "1 km", 
    "1 km", 
    "1 km"
  )
)


# Render the table using kable and kableExtra for a polished look
data_structure %>%
  kable("html", escape = FALSE, col.names = c("Indicator", "Description", "Source", "Spatial Resolution")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# Export the table to a CSV file
write.csv(data_structure, here("outputs", "tables", "summary_indicators.csv"), row.names = FALSE)
```


## Plot showing temporal change landscape
(Land cover, forest fragmentation, fire and fire pollutants.

```{r}
# color gradient for each country, from darkest to lightest
color_palette <- c("#3E2723", "#5D4037", "#795548", "#8D6E63", "#A1887F", "#BCAAA4", "#D7CCC8")

country_order <- c("Brasil", "Peru", "Bolivia", "Colombia", "Venezuela", "Suriname", "Ecuador")
color_mapping <- data.frame(country = country_order, color_code = color_palette)
```

Calculate mean temporal trends
```{r}
data_with_colors <- combined_data %>%
  dplyr::left_join(color_mapping, by = "country") %>%
  group_by(country, year) %>%
  summarize(
    for_PLAND = mean(for_PLAND, na.rm = TRUE),
    for_PD = mean(for_PD, na.rm = TRUE),
    for_ED = mean(for_ED, na.rm = TRUE),
    for_AI = mean(for_AI, na.rm = TRUE),
    savanna_PLAND = mean(savanna_PLAND, na.rm = TRUE),
    notforest_PLAND = mean(notforest_PLAND, na.rm = TRUE),
    color_code = unique(color_code)  # Retain color code after grouping
  ) %>%
  ungroup() %>%
  mutate(
    country = case_when(
      country == "Bolivia" ~ "BO",
      country == "Brasil" ~ "BR",
      country == "Colombia" ~ "CO",
      country == "Ecuador" ~ "EC",
      country == "Guiana FR" ~ "GUF",
      country == "Peru" ~ "PE",
      country == "Suriname" ~ "SR",
      country == "Venezuela" ~ "VE",
      TRUE ~ country  # Keep original value for unmatched cases
    ))

# Find the last year for each country to place the labels
label_data <- data_with_colors %>%
  group_by(country) %>%
  filter(year == max(year)) %>%
  ungroup()
```


### Forest Cover (for_PLAND)

```{r}
# Plotting with ribbons for standard deviation and country labels
(plot_forest_cover <- ggplot(data_with_colors, aes(x = year, y = for_PLAND, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
  scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
  breaks = seq(2000, 2020, by = 1),  # Ensure all years have a tick
  labels = seq(2000, 2020, by = 1),  # Explicitly define labels for all ticks
  expand = c(0, 0)  # No padding on the x-axis
) +
  scale_y_continuous(
    limits = c(0, 100),  # Fix y-axis range from 0 to 100
    expand = c(0, 0)  # No padding on the y-axis
  ) +
  labs(
    title = "Forest Cover Percent",
    x = " ",
    y = " "
  ) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),  # Increase x-axis label size
    axis.text.y = element_text(size = 16),  # Increase y-axis label size
    axis.title.x = element_text(size = 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size = 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size = 16),  # Increase legend text size
    legend.title = element_text(size = 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = for_PLAND, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size = 5
  ) +
  xlim(2000, 2020))
```

### Savanna Cover (savanna_PLAND)

```{r}
# Plotting with ribbons for standard deviation and country labels
(plot_savanna_cover <- ggplot(data_with_colors, aes(x = year, y = savanna_PLAND, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
  scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
  breaks = seq(2000, 2020, by = 1),  # Ensure all years have a tick
  labels = seq(2000, 2020, by = 1),  # Explicitly define labels for all ticks
  expand = c(0, 0)  # No padding on the x-axis
) +
  scale_y_continuous(
    limits = c(-0.5, 5),  # Fix y-axis range from 0 to 100
    expand = c(0, 0)  # No padding on the y-axis
  ) +
  labs(
    title = "Savanna Cover Percent",
    x = " ",
    y = " "
  ) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),  # Increase x-axis label size
    axis.text.y = element_text(size = 16),  # Increase y-axis label size
    axis.title.x = element_text(size = 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size = 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size = 16),  # Increase legend text size
    legend.title = element_text(size = 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = savanna_PLAND, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size = 5
  ) +
  xlim(2000, 2020))
```
### Non-forest Cover (notforest_PLAND)

```{r}
# Plotting with ribbons for standard deviation and country labels
(plot_notforest_cover <- ggplot(data_with_colors, aes(x = year, y = notforest_PLAND, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
  scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
  breaks = seq(2000, 2020, by = 1),  # Ensure all years have a tick
  labels = seq(2000, 2020, by = 1),  # Explicitly define labels for all ticks
  expand = c(0, 0)  # No padding on the x-axis
) +
  scale_y_continuous(
    limits = c(0, 40),  # Fix y-axis range from 0 to 100
    expand = c(0, 0)  # No padding on the y-axis
  ) +
  labs(
    title = "Non-forest Cover %",
    x = " ",
    y = " "
  ) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),  # Increase x-axis label size
    axis.text.y = element_text(size = 16),  # Increase y-axis label size
    axis.title.x = element_text(size = 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size = 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size = 16),  # Increase legend text size
    legend.title = element_text(size = 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = notforest_PLAND, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size = 5
  ) +
  xlim(2000, 2020))
```

### Patch Density (for_PD)
```{r}
(plot_patch_density <- ggplot(data_with_colors, aes(x = year, y = for_PD, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
   scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
    breaks = seq(2000, 2020, by = 1),
    labels = seq(2000, 2020, by = 1),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    limits = c(0, NA),  # Start from 0, no upper limit
    expand = c(0, 0)
  ) +
  labs(
    title = "Patch Density",
    x = " ",
    y = " "
  ) +
  theme_cowplot() +
  
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size= 16),  # Increase x-axis label size
    axis.text.y = element_text(size= 16),  # Increase y-axis label size
    axis.title.x = element_text(size= 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size= 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size= 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size= 16),  # Increase legend text size
    legend.title = element_text(size= 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = for_PD, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size= 5
  ) +
  xlim(2000, 2020))
```
### Edge Density (for_ED)
```{r}
(plot_edge_density <- ggplot(data_with_colors, aes(x = year, y = for_ED, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
  # geom_ribbon(
  #   aes(
  #     ymin = pmax(for_ED_mean - for_ED_sd, 0),  # Ensure ymin is at least 0
  #     ymax = for_ED_mean + for_ED_sd,  # No upper limit for edge density
  #     fill = color_code
  #   ),
  #   alpha = 0.1, color = NA
  # ) +
  scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
    breaks = seq(2000, 2020, by = 1),
    labels = seq(2000, 2020, by = 1),
    expand = c(0, 0)
  ) +
  # scale_y_continuous(
  #   limits = c(0, NA),  # Start from 0, no upper limit
  #   expand = c(0, 0)
  # ) +
  labs(
    title = "Edge Density",
    x = "Year",
    y = " "
  ) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size= 16),  # Increase x-axis label size
    axis.text.y = element_text(size= 16),  # Increase y-axis label size
    axis.title.x = element_text(size= 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size= 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size= 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size= 16),  # Increase legend text size
    legend.title = element_text(size= 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = for_ED, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size= 5
  ) +
  xlim(2000, 2020))
```

### Aggregation Index (for_AI)
```{r}
(plot_aggregation <- ggplot(data_with_colors, aes(x = year, y = for_AI, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
   scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
    breaks = seq(2000, 2020, by = 1),
    labels = seq(2000, 2020, by = 1),
    expand = c(0, 0)
  ) +
  labs(
    title = "Aggregation Index",
    x = " ",
    y = " "
  ) +
  # theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size= 16),  # Increase x-axis label size
    axis.text.y = element_text(size= 16),  # Increase y-axis label size
    axis.title.x = element_text(size= 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size= 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size= 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size= 16),  # Increase legend text size
    legend.title = element_text(size= 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = for_AI, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size= 5
  ) +
  xlim(2000, 2020))
```
## Plot panel landscape variables

```{r, fig.width = 15, fig.height = 9}
# Create the panel layout with your original positioning
panel_plot <- (plot_forest_cover | plot_savanna_cover | plot_notforest_cover) /
              (plot_patch_density | plot_edge_density | plot_aggregation) +
    plot_annotation(tag_levels = 'a')


panel_plot

ggsave(here("outputs", "figures", "landscape_change.jpg"), width = 18, height = 11)
```


### Fire Pollutant  (pm25_SUM)

Be attentive to Suriname, in which PM values are very high, also the only at country level or might be due to winds in the region - check.

```{r}
data_with_colors <- combined_data %>%
  left_join(color_mapping, by = "country") %>%
  group_by(country, year) %>%
  summarise(
    pm25_SUM_mean = mean(pm25_SUM, na.rm = TRUE),
    pm25_SUM_sd = sd(pm25_SUM, na.rm = TRUE),
    color_code = unique(color_code)
  ) %>%
  ungroup() %>%
  mutate(
    country = case_when(
      country == "Bolivia" ~ "BO",
      country == "Brasil" ~ "BR",
      country == "Colombia" ~ "CO",
      country == "Ecuador" ~ "EC",
      country == "Guiana FR" ~ "GUF",
      country == "Peru" ~ "PE",
      country == "Suriname" ~ "SR",
      country == "Venezuela" ~ "VE",
      TRUE ~ country  # Keep original value for unmatched cases
    ),
    pm25_SUM_mean = ifelse(is.nan(pm25_SUM_mean), NA, pm25_SUM_mean),
    pm25_SUM_sd = ifelse(is.nan(pm25_SUM_sd), NA, pm25_SUM_sd)
  ) %>% 
  drop_na(pm25_SUM_mean)

label_data <- data_with_colors %>%
  group_by(country) %>%
  filter(year == max(year)) %>%
  ungroup()

(plot_pm25 <- ggplot(data_with_colors, aes(x = year, y = pm25_SUM_mean, group = country)) +
  geom_line(aes(color = color_code), size = 1) +
  # geom_ribbon(
  #   aes(
  #     ymin = pmax(pm25_SUM_mean - pm25_SUM_sd, 0),  # Ensure ymin is at least 0
  #     ymax = pm25_SUM_mean + pm25_SUM_sd,  # No upper limit for patch density
  #     fill = color_code
  #   ),
  #   alpha = 0.1, color = NA
  # ) +
  scale_color_identity() +
  scale_fill_identity() +
  scale_x_continuous(
    breaks = seq(2000, 2020, by = 1),
    labels = seq(2000, 2020, by = 1),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    # limits = c(0, 100000000000),  # Start from 0, no upper limit
    expand = c(0, 0)
  ) +
  labs(
    # title = "Fire Pollutant (PM2.5) Over Years by Country",
    x = "Year",
    y = expression("Fire Pollutant (PM"[2.5]*")")
  ) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size= 16),  # Increase x-axis label size
    axis.text.y = element_text(size= 16),  # Increase y-axis label size
    axis.title.x = element_text(size= 16, face = "bold"),  # Increase x-axis title size
    axis.title.y = element_text(size= 16, face = "bold"),  # Increase y-axis title size
    plot.title = element_text(size= 16, face = "bold", hjust = 0.5),  # Increase and center title size
    legend.text = element_text(size= 16),  # Increase legend text size
    legend.title = element_text(size= 16)  # Increase legend title size
  ) +
  geom_text(
    data = label_data,
    aes(x = year, y = pm25_SUM_mean, label = country, color = color_code),
    hjust = -0.2,
    vjust = 0.5,
    size= 5
  ) +
  xlim(2000, 2020))
```




